A Hierarchical Bayesian Framework for Inferring the Stellar Obliquity Distribution

### 2023-03-21 src/tex/ms.tex
[showyourwork] overleaf sync

\newcommand{\dfm}[1]{\textcolor{blue}{\textbf{From DFM:} #1}}
\newcommand{\edits}[1]{\textcolor{grey}{\textbf{Edits:} #1}}

@@

For the stellar obliquity distribution, we model the $\cos{\psi}$ distribution instead of $\psi$ distribution to understand if the stellar obliquity is isotropically distributed.
If stellar obliquity distribution is isotropic, $\cos{\psi}$ is uniformly distributed between -1 to 1.
To flexibly model the stellar obliquity distribution, we select a two-component mixture of Betas distribution with hyperparameters $\bm{\beta} = \{\bm{w},\bm{a},\bm{b}\}$, where each hyperparameter has a dimension of 2.
This mixture distribution has the capacity to capture anything from an isotropic distribution to a strongly bimodal population.
% One component is designed for a large population of well-aligned systems, and the other is designed for misaligned systems. The Beta distribution is a flexible distribution that can describe the spike of aligned systems with stellar obliquities close to $0\degr$ or the clustered or broadly distributed misaligned systems.
\dfm{The notation here is a bit sloppy. It might be better to say something like: there's some parameter $x \sim \mathrm{mixture}$ which we then transform as $2\,x-1$, which isn't quite the same.}
Then, under this model, the probability density function for $\cos{\psi}$ is 
\begin{align}
    \cos{\psi} &\sim 2\,\Bigl( w_0 {\rm Beta}(a_0, b_0) + w_1 {\rm Beta}(a_1, b_1)\Bigr)-1.
\end{align}
Since the Beta distribution is defined on the interval $[0, 1]$, we extend its support to $[-1,1]$ using a linear transformation (i.e., $2\,{\rm Beta} - 1$).

\dfm{I'm not sure that I understand this next bit. Why and how do we do this?}
For the ${\rm Beta}(a_0,b_0)$ component, we have it to describe the population of aligned systems. For the $a_0/b_0 \gg 1$, the distribution will have a spike at $\cos{\psi} = 1$. For the ${\rm Beta}(a_1,b_1)$ component, we have it to describe the misaligned systems.

We design the order of the two components to prevent label switching in the mixture model \dfm{how?}.
With the design, the hyperparameter and parameter priors are the following:
\begin{align}
    w_{0,1} &\sim {\rm Dirichlet}(1, 1) \nonumber\\
    a_0 &\sim \mathcal{U}(0, 50) \nonumber\\
    b_0 &\sim \mathcal{U}(0, 1) \nonumber\\
    a_1 &\sim \mathcal{U}(0, 10) \nonumber\\
    b_1 &\sim \mathcal{U}(0, 10) \nonumber\\
    \cos{i}_{\star,n} &\sim \mathcal{U}(0, 1) \nonumber\\
    \cos{i}_{{\rm orb},n} &\sim \mathcal{U}(0, 1) \nonumber\\
    \lambda_{\star,n} &\sim \mathcal{U}(-\pi, \pi).
\end{align}
The likelihood functions follow:
\begin{align}
    \mathcal{L}(\lambda) &\sim \prod_{i=1}^N\mathcal{N}(\hat{\lambda}_n, \sigma_{\hat{\lambda}_n}) \nonumber\\
    \mathcal{L}(i_{\rm orb}) &\sim \prod_{i=1}^N\mathcal{N}(\hat{i}_{{\rm orb},n}, \sigma_{\hat{i}_{{\rm orb},n}}) \nonumber\\
    \mathcal{L}(i_\star) &\sim \prod_{i=1}^N\mathcal{N}({\rm Obs}_{\star,n}, \sigma_{{\rm Obs}_{\star,n}}).
\end{align}
If $\gamma_{\star, n}$ is available, we construct Normal distributions with means and standard deviations from measurements.
Here we use uniform hyperpriors for the Beta distribution. However, it is important to note that when applying the framework to a small sample size with $N \lesssim 50$, the choice of hyperpriors could impact the inferred distribution \citep{Nagpal22}. To ensure the robustness of the inferred distributions in such cases, it is crucial to test their sensitivity to different hyperpriors.

@@

In Figure~\ref{fig:simulation}, we present the results of this experiment, plotting the inferred stellar obliquity distributions, compared to the ground-truth distributions.
Since the simulated stellar obliquity distributions only have a single component, we model the data with a single Beta distribution.
Each row of Figure~\ref{fig:simulation} corresponds to a different simulation distribution. The orange curve and contours in the left column are the median, 1-$\sigma$ and 2-$\sigma$ uncertainties of the inferred $\cos{\psi}$ distribution with stellar inclination information.
The right column shows the same inferences (in blue) when constraints on stellar inclination are not included.
% We simply assume an isotropic distribution of $i_\star$ for these models.
Surprisingly, Figure~\ref{fig:simulation} demonstrates that our inference procedure recovers the true distribution for $\cos{\psi}$ equally well, regardless of the inclusion of $i_\star$ measurements.
Despite the fact that the inferred distributions without $i_\star$ measurements have marginally wider uncertainties, as indicated by shallow color contours, the modes and widths of the inferred stellar obliquity distributions are consistent with or without $i_\star$ likelihood.
Since the injected distributions in rows 2-4 are Normal distributions, it should not be surprising that the inferred distributions, which are Beta distributions, may not exactly match the injected distributions.
\dfm{Add some more here about how the true distribution is not included within the support of the distribution we're using to fit. One thought: perhaps we would actually do better if we used the two betas here too?}

@@

The $\{\psi, \theta\}$ coordinate system shown in panel (a) relates to the physical properties of a planetary system. $\psi$ is the angle between the stellar spin axis and the planetary orbital axis, and $\theta$ is the azimuthal angle of the stellar spin axis. For the given $\bf{n}_{\rm orb}$ axis, if $\bf{n}_{\star}$ is a random vector with uniform distribution on a three-dimensional sphere, i.e., $\bf{n}_{\star}$ is isotropically distributed, the probability density function of $\psi$ follows $p_{\psi} \sim \sin{\psi}$ and $p_\theta \sim 1/2\pi$. Since $p_{\cos{\psi}} = p_{\psi} \abs{\dv*{\psi}{\cos{\psi}}}$, $p_{\cos{\psi}}$ is uniformly distributed between $-1$ and $1$ for isotropic stellar obliquity.
The $\{\lambda, i_\star\}$ coordinate system shown in panel (b) relates to observed properties. $\lambda$ is the sky-projected stellar obliquity and $i_\star$ is the stellar inclination. If $\bf{n}_{\star}$ is isotropically distributed, $p_{i_\star} \sim \sin{i_\star}$ and $p_\lambda \sim \mathcal{U}(-\pi, \pi)$. Again, since $p_{\cos{i_\star}} = p_{i_\star} \abs{\dv*{i_\star}{\cos{i_\star}}}$, $p_{\cos{i_\star}}$ is uniformly distributed between $-1$ and $1$. 
Because of the observational degeneracy between $i_\star$ and $180\degr - i_\star$, the convention is to have $0 \leqslant i_\star \leqslant 90\degr$, and thus $p_{\cos{i_\star}}$ is uniformly distributed between $0$ and $1$ and $p_{\cos{i_\star}} \sim \mathcal{U}(0, 1)$. This also limits $p_\theta \sim \mathcal{U}(-\pi/2, \pi/2)$.

@@

First, we derive the $\lambda$ distribution for a given $\cos{\psi}$ distribution. We could find the distribution of $\cos{\lambda}$ using the Jacobian transformation from $\cos{\psi}$ and $\cos{\theta}$. Since $\psi$ and $\theta$ are independent variables, we could marginalize over $\theta$ to find the relation between the probability density functions between $\lambda$ and $\psi$.
The Jacobian transformation follows
\begin{equation}
    p_{\cos{\lambda}} = \int \abs{\pdv{\cos{\psi}}{\cos{\lambda}}} p_{\cos{\psi}} p_{\cos{\theta}} d\cos{\theta}.
\end{equation}
Replacing $\sin{i_\star}$ in Equation~(\ref{eq:coord3}) using Equation~(\ref{eq:coord1}), we find $\cos{\lambda} = \cos{\psi}/\sqrt{1-(1-\cos^2{\psi})\cos^2{\theta}}$. Reorganize the equation, we get $\cos^2{\psi} = \frac{\cos^2{\lambda}\cos^2{\theta}-\cos^2{\lambda}}{\cos^2{\lambda}\cos^2{\theta}-1}$ and
the partial derivative $\abs{\pdv{\cos{\psi}}{\cos{\lambda}}} = \frac{(1-\cos^2{\theta})^{1/2}}{(1-\cos^2{\theta}\cos^2{\lambda})^{3/2}}$. Assume $\theta$ is uniformly distributed between $-\pi/2$ and $\pi/2$, $p_{\cos{\theta}} = p_{\theta} \abs{\dv*{\theta}{\cos{\theta}}} = 2/\pi/(1-\cos^2{\theta)^{1/2}}$, where the factor of 2 is due to two solutions of $\theta$ to the $\cos{\theta}$. Putting all the parts together, we get
\begin{equation}\label{eqn:jac_lam}
    p_{\cos{\lambda}} = \frac{2}{\pi} \int_{0}^{1} (1-\cos^2{\theta}\cos^2{\lambda})^{-3/2} p_{\cos{\psi}} d\cos{\theta}.
\end{equation}
If $\cos{\psi}$ is uniformly distribution, i.e., $p_{\cos{\psi}} \sim 1/2$, Equation~(\ref{eqn:jac_lam}) becomes $p_{\cos{\lambda}} \sim 1/\pi/\sqrt{1-\cos^2{\lambda}}$, which is equivalent to $p_\lambda \sim \mathcal{U}(-\pi, \pi)$. This suggests $\lambda$ is uniformly distributed for an isotropic $\psi$ distribution. For different $\cos{\psi}$ distributions, $p_\lambda$ can be found by evaluating Equation~(\ref{eqn:jac_lam}).

Next, we derive the $i_\star$ distribution for a given $\cos{\psi}$ distribution. Similarly, we first find the Jacobian transformation of $i_\star$ from $\psi$ and $\theta$ and then marginalize over $\theta$. It is easier to work on $\cos{i_\star}$ than $i_\star$:
\begin{equation}
    p_{\cos{i_\star}} = \int \abs{\pdv{\sin{\psi}}{\cos{i_\star}}} p_{\sin{\psi}} p_{\cos{\theta}} d\cos{\theta}.
\end{equation}
From Equation~(\ref{eq:coord1}), we have $\sin{\psi} = \cos{i_\star}/\cos{\theta}$ and $\abs{\pdv{\sin{\psi}}{\cos{i_\star}}} = 1/\cos{\theta}$. Again, we assume $\theta$ is uniformly distributed, and this gives $p_{\cos{\theta}} = 2/\pi/(1-\cos^2{\theta)^{1/2}}$. Lastly, we transform the $p_{\sin{\psi}}$ to $p_{\cos{\psi}}$, where $p_{\sin{\psi}} = 2p_{\cos{\psi}}\sin{\psi}/\sqrt{1-\sin^2{\psi}}$ and the factor of 2 is from two solutions of $\cos{\psi}$ to the $\cos^2{\psi} = 1-\sin^2{\psi}$. Combining all the pieces together, we get
\begin{equation}\label{eqn:jac_istar}
    p_{\cos{i_\star}} = \frac{4}{\pi} \int_{\cos{i_\star}}^{1} \frac{\cos{i_\star}/\cos{\theta}}{\sqrt{\cos^2{\theta}-\cos^2{i_\star}}} \frac{1}{\sqrt{1-\cos^2{\theta}}} p_{\cos{\psi}} d\cos{\theta}.
\end{equation}
Note that the lower limit of the integral is $\cos{i_\star}$ instead of 0 since $\lvert \cos{\theta}/\cos{i_\star} \rvert \geqslant 1$.
If $\cos{\psi}$ is uniformly distribution, i.e., $p_{\cos{\psi}} \sim 1/2$, the integral gives $1$, which suggests the $\cos{i_\star}$ is uniformly distributed, as expected.

>>>>

\newcommand{\dfm}[1]{\textcolor{blue}{\textbf{From DFM:} #1}}
\newcommand{\edits}[1]{\textcolor{purple}{#1}}

@@

For the stellar obliquity distribution, we model the $\cos{\psi}$ distribution instead of $\psi$ distribution to understand if the stellar obliquity is isotropically distributed.
If stellar obliquity distribution is isotropic, $\cos{\psi}$ is uniformly distributed between -1 and 1.
To flexibly model the stellar obliquity distribution, we select a two-component mixture of Betas distribution with hyperparameters $\bm{\beta} = \{\bm{w},\bm{a},\bm{b}\}$, where each hyperparameter has a dimension of 2. \edits{We have $\bm{w} = \{w_0, w_1\}$, $\bm{a} = \{a_0, a_1\}$, and  $\bm{b} = \{b_0, b_1\}$.}
This mixture distribution has the capacity to capture anything from an isotropic distribution to a strongly bimodal population.
% One component is designed for a large population of well-aligned systems, and the other is designed for misaligned systems. The Beta distribution is a flexible distribution that can describe the spike of aligned systems with stellar obliquities close to $0\degr$ or the clustered or broadly distributed misaligned systems.
\dfm{The notation here is a bit sloppy. It might be better to say something like: there's some parameter $x \sim \mathrm{mixture}$ which we then transform as $2\,x-1$, which isn't quite the same.}
Then, under this model, the probability density function for $\cos{\psi}$ is 
\begin{align}
    u_0 &\sim {\rm Beta}(a_0, b_0) \nonumber\\
    u_1 &\sim {\rm Beta}(a_1, b_1) \nonumber\\
    \cos{\psi} &\sim 2\,( w_0 u_0 + w_1 u_1)-1.
\end{align}
Since the Beta distribution is defined on the interval $[0, 1]$, we extend its support to $[-1,1]$ using a linear transformation (i.e., $2\,( w_0 u_0 + w_1 u_1)-1$).
\dfm{I'm not sure that I understand this next bit. Why and how do we do this?}
For the ${\rm Beta}(a_0,b_0)$ component, we have it to describe the population of aligned systems. For the $a_0/b_0 \gg 1$, the distribution will have a spike at $\cos{\psi} = 1$. For the ${\rm Beta}(a_1,b_1)$ component, we have it to describe the misaligned systems.
% add more description on why chose a0, b0 priors
We design the order of the two components to prevent label switching in the mixture model \dfm{how?}.

With the design, the hyperparameter and parameter priors are the following:
\begin{align}
    w_{0,1} &\sim {\rm Dirichlet}(1, 1) \nonumber\\
    a_0 &\sim \mathcal{U}(0, 50) \nonumber\\
    b_0 &\sim \mathcal{U}(0, 1) \nonumber\\
    a_1 &\sim \mathcal{U}(0, 10) \nonumber\\
    b_1 &\sim \mathcal{U}(0, 10) \nonumber\\
    \cos{i}_{\star,n} &\sim \mathcal{U}(0, 1) \nonumber\\
    \cos{i}_{{\rm orb},n} &\sim \mathcal{U}(0, 1) \nonumber\\
    \lambda_{\star,n} &\sim \mathcal{U}(-\pi, \pi).
\end{align}
The likelihood functions follow:
\begin{align}
    \mathcal{L}(\lambda) &\sim \prod_{i=1}^N\mathcal{N}(\hat{\lambda}_n, \sigma_{\hat{\lambda}_n}) \nonumber\\
    \mathcal{L}(i_{\rm orb}) &\sim \prod_{i=1}^N\mathcal{N}(\hat{i}_{{\rm orb},n}, \sigma_{\hat{i}_{{\rm orb},n}}) \nonumber\\
    \mathcal{L}(i_\star) &\sim \prod_{i=1}^N\mathcal{N}({\rm Obs}_{\star,n}, \sigma_{{\rm Obs}_{\star,n}}).
\end{align}
If $\gamma_{\star, n}$ is available, we construct Normal distributions with means and standard deviations from measurements.
Here we use uniform hyperpriors for the Beta distribution. However, it is important to note that when applying the framework to a small sample size with $N \lesssim 50$, the choice of hyperpriors could impact the inferred distribution \citep{Nagpal22}. To ensure the robustness of the inferred distributions in such cases, it is crucial to test their sensitivity to different hyperpriors.

@@

In Figure~\ref{fig:simulation}, we present the results of this experiment, plotting the inferred stellar obliquity distributions, compared to the ground-truth distributions.
Since the simulated stellar obliquity distributions only have a single component, we model the data with a single Beta distribution.
Each row of Figure~\ref{fig:simulation} corresponds to a different simulation distribution. The orange curve and contours in the left column are the median, 1-$\sigma$ and 2-$\sigma$ uncertainties of the inferred $\cos{\psi}$ distribution with stellar inclination information.
The right column shows the same inferences (in blue) when constraints on stellar inclination are not included.
Surprisingly, Figure~\ref{fig:simulation} demonstrates that our inference procedure recovers the true distribution for $\cos{\psi}$ equally well, regardless of the inclusion of $i_\star$ measurements.
Despite the fact that the inferred distributions without $i_\star$ measurements have marginally wider uncertainties, as indicated by shallow color contours, the modes and widths of the inferred stellar obliquity distributions are consistent with or without $i_\star$ likelihood.
Since the injected distributions in rows 2-4 are Normal distributions, it should not be surprising that the inferred distributions, which are Beta distributions, may not exactly match the injected distributions.
\dfm{Add some more here about how the true distribution is not included within the support of the distribution we're using to fit. One thought: perhaps we would actually do better if we used the two betas here too?}

@@

The $\{\psi, \theta\}$ coordinate system shown in panel (a) relates to the physical properties of a planetary system. $\psi$ is the angle between the stellar spin axis and the planetary orbital axis, and $\theta$ is the azimuthal angle of the stellar spin axis. For the given $\bf{n}_{\rm orb}$ axis, if $\bf{n}_{\star}$ is a random vector with uniform distribution on a three-dimensional sphere, i.e., $\bf{n}_{\star}$ is isotropically distributed, the probability density function of $\psi$ follows $p(\psi) = \sin{\psi}$ and $p(\theta) = 1/2\pi$. 
\edits{Here $p(x)$ denote the probability density distribution of $x$.}
Since $p(\cos{\psi}) = p(\psi) \abs{\dv*{\psi}{\cos{\psi}}}$, $p(\cos{\psi})$ is uniformly distributed between $-1$ and $1$ for isotropic stellar obliquity.
The $\{\lambda, i_\star\}$ coordinate system shown in panel (b) relates to observed properties. $\lambda$ is the sky-projected stellar obliquity and $i_\star$ is the stellar inclination. If $\bf{n}_{\star}$ is isotropically distributed, $p(i_\star) = \sin{i_\star}$ and $p(\lambda) = 1/2\pi$. Again, since $p(\cos{i_\star}) = p(i_\star) \abs{\dv*{i_\star}{\cos{i_\star}}}$, $p(\cos{i_\star})$ is uniformly distributed between $-1$ and $1$. 
Because of the observational degeneracy between $i_\star$ and $\pi - i_\star$, the convention is to have $0 \leqslant i_\star \leqslant \pi/2$, and thus $\cos{i_\star} \sim \mathcal{U}(0, 1)$ (i.e., $p(\cos{i_\star}) = 1$). This also suggests $\theta \sim \mathcal{U}(-\pi/2, \pi/2)$.

@@

First, we derive the $\lambda$ distribution for a given $\cos{\psi}$ distribution. We could find the distribution of $\cos{\lambda}$ using the Jacobian transformation from $\cos{\psi}$ and $\cos{\theta}$. Since $\psi$ and $\theta$ are independent variables, we could marginalize over $\theta$ to find the relation between the probability density functions between $\lambda$ and $\psi$.

The Jacobian transformation follows
\begin{equation}
    p(\cos{\lambda}) = \int \abs{\pdv{\cos{\psi}}{\cos{\lambda}}} p(\cos{\psi}) p(\cos{\theta}) d\cos{\theta}.
\end{equation}
Replacing $\sin{i_\star}$ in Equation~(\ref{eq:coord3}) using Equation~(\ref{eq:coord1}), we find 
\begin{equation}
    \cos{\lambda} = \cos{\psi}/\sqrt{1-(1-\cos^2{\psi})\cos^2{\theta}}.
\end{equation} 
Reorganize the equation, we get 
\begin{equation}
    \cos^2{\psi} = \frac{\cos^2{\lambda}\cos^2{\theta}-\cos^2{\lambda}}{\cos^2{\lambda}\cos^2{\theta}-1}
\end{equation}
and the partial derivative 
\begin{equation}
    \abs{\pdv{\cos{\psi}}{\cos{\lambda}}} = \frac{(1-\cos^2{\theta})^{1/2}}{(1-\cos^2{\theta}\cos^2{\lambda})^{3/2}}.
\end{equation}
Assume $\theta$ is uniformly distributed between $-\pi/2$ and $\pi/2$, $p(\cos{\theta}) = p(\theta) \abs{\dv*{\theta}{\cos{\theta}}} = 2/\pi/(1-\cos^2{\theta)^{1/2}}$, where the factor of 2 is due to two solutions of $\theta$ to the $\cos{\theta}$. Putting all the parts together, we get
\begin{equation}\label{eqn:jac_lam}
    p(\cos{\lambda}) = \frac{2}{\pi} \int_{0}^{1} (1-\cos^2{\theta}\cos^2{\lambda})^{-3/2} p(\cos{\psi}) d\cos{\theta}.
\end{equation}
If $\cos{\psi}$ is uniformly distribution, i.e., $p(\cos{\psi}) = 1/2$, Equation~(\ref{eqn:jac_lam}) becomes $p(\cos{\lambda}) = 1/\pi/\sqrt{1-\cos^2{\lambda}}$, which is equivalent to $\lambda \sim \mathcal{U}(-\pi, \pi)$. This suggests $\lambda$ is uniformly distributed for an isotropic $\psi$ distribution. For different $\cos{\psi}$ distributions, $p(\lambda)$ can be found by evaluating Equation~(\ref{eqn:jac_lam}).

Next, we derive the $i_\star$ distribution for a given $\cos{\psi}$ distribution. Similarly, we first find the Jacobian transformation of $i_\star$ from $\psi$ and $\theta$ and then marginalize over $\theta$. It is easier to work on $\cos{i_\star}$ than $i_\star$:
\begin{equation}
    p(\cos{i_\star}) = \int \abs{\pdv{\sin{\psi}}{\cos{i_\star}}} p(\sin{\psi}) p(\cos{\theta}) d\cos{\theta}.
\end{equation}
From Equation~(\ref{eq:coord1}), we get 
\begin{equation}
    \sin{\psi} = \cos{i_\star}/\cos{\theta}
\end{equation} and the partial derivative
\begin{equation}
    \abs{\pdv{\sin{\psi}}{\cos{i_\star}}} = 1/\cos{\theta}.
\end{equation} 
Again, we assume $\theta$ is uniformly distributed, and this gives $p(\cos{\theta}) = 2/\pi/(1-\cos^2{\theta)^{1/2}}$. Lastly, we transform the $p(\sin{\psi})$ to $p(\cos{\psi})$,
\begin{equation}
    p(\sin{\psi}) = 2p(\cos{\psi})\sin{\psi}/\sqrt{1-\sin^2{\psi}}.
\end{equation}
The factor of 2 is from two solutions of $\cos{\psi}$ to the $\cos^2{\psi} = 1-\sin^2{\psi}$. Combining all the pieces together, we get
\begin{align}\label{eqn:jac_istar}
    p(\cos{i_\star}) = \frac{4}{\pi} \int_{\cos{i_\star}}^{1}& \frac{\cos{i_\star}/\cos{\theta}}{\sqrt{\cos^2{\theta}-\cos^2{i_\star}}} \nonumber\\ 
    &\frac{1}{\sqrt{1-\cos^2{\theta}}} p(\cos{\psi}) d\cos{\theta}.
\end{align}
Note that the lower limit of the integral is $\cos{i_\star}$ instead of 0 since $\lvert \cos{\theta}/\cos{i_\star} \rvert \geqslant 1$.
If $\cos{\psi}$ is uniformly distribution, i.e., $p(\cos{\psi}) = 1/2$, the integral gives $1$, which suggests the $\cos{i_\star}$ is uniformly distributed, as expected.
